|

### 2025-03-23 reproducibility/manuscript/manuscript.qmd
fix(manuscript): make model formulation eqs align
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
\begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1.
  \end{cases}
\end{align}

Now, we describe the priors for the initial conditions, noting that this is the only difference between Model 1 and Model 2:
\begin{align}
  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
    (0, 0) & \text{Model 1} \\
    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
  \end{cases} \\
  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
  \end{cases}
\end{align}

We define the ODE solution at time $\tau_{cg}$ as:
\begin{equation}
  \label{eq-ODE-solve}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
\end{equation}

@@

The linear ODE in Eq. \ref{eq-rna} has an analytic solution under the assumption that the transcription rate $\alpha(t)$ is piece-wise constant, and is derived by the authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa].
For brevity, we present the solution under constant transcription rate $\alpha(t) = \alpha_0$, and note that this can be combined to assemble the solution in the piece-wise constant case.
First, let $C = u_0 - \frac{\alpha}{\beta}$. Then, the solution to the ODE in Eq. \ref{eq-rna} is:
\begin{equation}
\label{eq-rna-solution}
\begin{aligned}
  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
  s(t) &= \begin{cases}
    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
  \end{cases}
\end{aligned}
\end{equation}
These solutions are the essential building block used to compute all cell-gene-specific ODE solutions for the unspliced and spliced expression levels in the generative model.

>>>>

Next we introduce the priors for the splicing parameters (where the activation rate $\alpha$ depends on the activation state $k_{cg}$ from above):
\begin{align*}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1.
  \end{cases}
\end{align*}

Now, we describe the priors for the initial conditions, noting that these define the difference between Model 1 and Model 2, given respectively
\begin{align*}
  &\left(\hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}\right) \sim 
  \begin{cases}
    (0, 0) \\
    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1))
  \end{cases}
\end{align*}
where
\begin{align*}
  &\left(u^{(0)}_{cg}, s^{(0)}_{cg}\right) = \\
  &\hskip 3pt \begin{cases}
    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \\
    \hskip 24pt T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
  \end{cases}
\end{align*}

We define the ODE solution at time $\tau_{cg}$ as:
\begin{align}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, & s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \nonumber \\
    & T_0=0, T_1=\tau_{cg} \Big). \label{eq-ODE-solve}
\end{align}

@@

The linear ODE in Eq. \ref{eq-rna} has an analytic solution under the assumption that the transcription rate $\alpha(t)$ is piece-wise constant, and is derived by the authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa].
For brevity, we present the solution under constant transcription rate $\alpha(t) = \alpha_0$, and note that this can be combined to assemble the solution in the piece-wise constant case.
First, let $C = u_0 - \frac{\alpha}{\beta}$. Then, the solution to the ODE in Eq. \ref{eq-rna} is:
\begin{equation}
\label{eq-rna-solution}
\begin{aligned}
  &u(t) = \frac{\alpha}{\beta} + C e^{-\beta t} \\
  &s(t) = 
  \begin{cases}
    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} \\
    \hskip 12pt + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
    \frac{\alpha}{\beta} \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
  \end{cases}
\end{aligned}
\end{equation}
These solutions are the essential building block used to compute all cell-gene-specific ODE solutions for the unspliced and spliced expression levels in the generative model.
